From noise to knowledge: how randomness generates novel phenomena and information

### 2017-10-27 appendixA/appendixA.Rmd
appendixA
WIP

---
title: "Appendix"
output: pdf_document
---

@@

## plot themes, non-essential
library(hrbrthemes)
library(ggthemes)
ggplot2::theme_set(hrbrthemes::theme_ipsum())
```

@@

```{r}
read_csv("quasicycles.csv") %>%
  gather(species, pop, -t, -sigma) %>%
  ggplot(aes(t, pop, col=species)) + geom_line() + facet_wrap(~sigma) + scale_color_ptol()
```

@@

May's model

```{r}
p <- list(r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 1e4)

x <- seq(0,2, length.out = 100)
data.frame(x, f = f(x), g = g(x)) %>%
  ggplot(aes(x)) + geom_line(aes(y = f-g)) + 
  geom_hline(aes(yintercept=0), lty=2) +
  coord_cartesian(xlim = c(0,1.5), ylim=c(-.015,.1))

>>>>

---
title: "Appendix A: Stochastic simulations of noisy phenomena"
output: pdf_document
---

@@

## plot themes, non-essential
library(hrbrthemes)
library(ggthemes)
ggplot2::theme_set(hrbrthemes::theme_ipsum())
colours <- ptol_pal()(2)

@@

```{r}
read_csv("quasicycles.csv") %>%
  gather(species, pop, -t, -sigma) %>%
  mutate(sigma = as.factor(sigma)) %>%
  mutate(sigma = recode(sigma, "1e-05" = "A.", "0.01" = "B.")) %>%
  ggplot(aes(t, pop, col=species)) +
  geom_line() + 
  facet_wrap(~sigma) + 
  scale_color_ptol()
```

@@

May's model:

$$X_{t+1} = X_t + \underbrace{X_t r \left(1 -\frac{X_t}{K} \right)}_{\textrm{Vegitation growth}}  - \underbrace{\frac{a X_t ^ Q}{X_t^ Q + H ^ Q}}_{\textrm{Vegitation consumption}} + \xi_t,$$

```{r}
p <- list(r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 1e4)
f <- function(x) x * p$r * (1 - x / p$K)
g <- function(x) p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)
x <- seq(0,2, length.out = 100)

p2 <- data.frame(x, f = f(x), g = g(x)) %>%
  ggplot(aes(x, y)) + 
  geom_line(aes(y = f), col=colours[1], lwd=1) +
  geom_line(aes(y = g), col=colours[2], lwd=1) + labs(x = bquote(X[t]), y = bquote(X[t+1]))

### 2017-10-27 paper.Rmd
WIP

keywords:
  - Stochasticity
  - Demographic noise
  - Environmental noise
  - Colored noise
  - quasi-cycles

@@

```{r libraries, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, dev="CairoPDF", fig.width=3.5, fig.height=3.5)
library(ggthemes)
library(tidyverse)
library(cowplot)
library(hrbrthemes)
extrafont::loadfonts()
#extrafont::font_import()                    # One-time use to set up fonts
#hrbrthemes::import_roboto_condensed()       # One-time use to set up fonts
ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
#ggplot2::theme_set(theme_bw()) # simpler theme if you do not have these fonts
```

@@

Approximations of the so-called master equation, van Kampen system size approximation [@vanKampen1987] described in @Black2010, (rooted in the theorems of @Kurtz1977),

@@

```{r warning=FALSE}
read_csv("appendix/gillespie.csv", col_types = "cdiddd")  %>% 
  ggplot(aes(x = time)) + 
  geom_hline(aes(yintercept = mean), lty=2) + 
  geom_hline(aes(yintercept = minus_sd), lty=2) + 
  geom_hline(aes(yintercept = plus_sd), lty=2) + 
  geom_line(aes(y = n), col="red") +
  facet_wrap(~system_size, scales = "free_y") + 
  labs(caption = 
"Population dynamics from a Gillespie simulation of the Levins model with large (N=1000) 
and small (N=100) number of sites (red) show relatively smaller effects of demographic 
noise in the larger systme. Models are otherwise identical, with e = 0.2 and c = 1 (code
in appendix A). Theoretical predictions for mean and plus/minus one standard deviation 
shown in horizontal black dashed lines.")
```

@@

**Quasi-cycles**

Understanding the origin of sustained oscillations has long been a central question in ecology [@Hastings1996; @Bjornstad2001].  The simple Lotka Volterra model produces only neutrally stable centers, while the more realistic model that introduces carrying capacity for prey as they compete for resources results in a model that produces only damped oscillations, eventually settling into a stable state (Figure 1a).  However, the introduction of relatively small noise into the system can be sufficient to excite the system into sustained oscillations (Figure 1b.)  In the example shown here, a simple predator-prey system is driven by environmental white noise $\xi_t$, with prey following the equation:

@@

$$y_{t+1} = y_t + c  x_t  y_t - d  y_t+ \xi_{y,t} $$
with $c$ the conversion ratio and $d$ the predator mortality rate.  Here we have assumed for the prey, $\xi_x \sim \mathcal{N(0, \sigma_x)}$, and similarly for the predator driven by $\xi_y$. Figure 1a shows examples for noise $\sigma_x = \sigma_y = 10^{-5}$, while panel 1b shows $\sigma_x = \sigma_y = 0.01$.

```{r quasicycles, fig.width=7}
read_csv("appendix/quasicycles.csv") %>%
  gather(species, population, -t, -sigma) %>%
  rename(time = t) %>%
  ggplot(aes(time, population, col = species)) + geom_line() + 
  facet_wrap(~ sigma, ncol=2) + scale_color_ptol() #+ 
#  labs(x="Time", 
#       y="Population density",
#       title="Noise induces quasi-cycles",
#       caption="Left panel: neglible  noise results only in damped oscillations.\nRight panel: small noise creates sustained oscillations.")
#ggsave("figures/quasicycles.pdf")
```

@@

**Stochastic switching**

We turn from stochastic resonance to oscillations of a very different sort, which do not resemble any deterministic process:

```{r noisy_switch}
read_csv("appendix/noisy_switch.csv")  %>% ggplot(aes(t,x)) + 
  geom_line() #+ 
  #labs(title="Stochastic switching", 
  #     subtitle = "Noise drives transitions between alternative stable states")
#ggsave("figures/noisy_switch.pdf")

```

@@

@@

```{r tipping, fig.width=7, fig.height=7}
tipping <- read_csv("appendix/tipping.csv")

p1 <- tipping %>%
  select(-x) %>%
  gather(series, value, -t, -tip_time) %>%
  ggplot(aes(t, value)) + geom_line() + 
  geom_vline(aes(xintercept = tip_time), col="red", lty=2) +
  facet_wrap(~series, ncol = 2, scales="free_y")  +
  theme_ipsum_rc() 
p2 <- tipping %>% 
  select(t, population = x, tip_time) %>%
  ggplot(aes(t, population)) + geom_line() + 
  geom_vline(aes(xintercept = tip_time), col="red", lty=2) + 
  #scale_y_log10()  + 
  theme_ipsum_rc() + 
  labs(subtitle = "Yeast population size", 
       caption = "Vertical red dashed line indicates tipping point location")

library(cowplot)
ggdraw() +
  draw_plot(p1, x = 0, y = .5, width = 1, height = 0.5) +
  draw_plot(p2, x = 0, y = 0, width = 1, height = 0.6)
#ggsave("figures/tipping.pdf")
```

>>>>

keywords:
  - stochasticity
  - demographic noise
  - environmental noise
  - colored noise
  - quasi-cycles

@@

```{r libraries, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, dev="CairoPDF", fig.width=7, fig.height=3.5)
library(ggthemes)
library(tidyverse)
library(cowplot)
library(hrbrthemes)
extrafont::loadfonts()
#extrafont::font_import()                    # One-time use to set up fonts
#hrbrthemes::import_roboto_condensed()       # One-time use to set up fonts
ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
#ggplot2::theme_set(theme_bw()) # simpler theme if you do not have these fonts
colours <- ptol_pal()(2)
```

@@

Approximations of the so-called master equation, van Kampen system size approximation [@vanKampen2007] described in @Black2010, (rooted in the theorems of [@Kurtz1970; @Kurtz1971]),

@@

<!-- Figure 1: Gillespie simulations -->
```{r gillespie, warning=FALSE, fig.width=7, fig.cap="Population dynamics from a Gillespie simulation of the Levins model with large (N=1000, panel A) and small (N=100, panel B) number of sites (blue) show relatively weaker effects of demographic noise in the bigger system. Models are otherwise identical, with e = 0.2 and c = 1 (code in appendix A). Theoretical predictions for mean and plus/minus one standard deviation shown in horizontal re dashed lines."}
read_csv("appendixA/gillespie.csv", col_types = "cdiddd")  %>% 
  mutate(system_size = recode(system_size, large = "A. 1000 total sites", small= "B. 100 total sites")) %>%
  ggplot(aes(x = time)) + 
  geom_hline(aes(yintercept = mean), lty=2, col=colours[[2]]) + 
  geom_hline(aes(yintercept = minus_sd), lty=2, col=colours[[2]]) + 
  geom_hline(aes(yintercept = plus_sd), lty=2, col=colours[[2]]) + 
  geom_line(aes(y = n), col=colours[[1]]) +
  facet_wrap(~system_size, scales = "free_y") 
```

@@

## Quasi-cycles

Understanding the origin of sustained oscillations has long been a central question in ecology [@Hastings1996; @Bjornstad2001].  The simple Lotka Volterra model produces only neutrally stable centers, while the more realistic model that introduces carrying capacity for prey as they compete for resources results in a model that produces only damped oscillations, eventually settling into a stable state (Figure 1A).  However, the introduction of relatively small noise into the system can be sufficient to excite the system into sustained oscillations (Figure 1B).  In the example shown here, a simple predator-prey system is driven by environmental white noise $\xi_t$, with prey following the equation:

@@

$$y_{t+1} = y_t + c  x_t  y_t - d  y_t+ \xi_{y,t} $$
with $c$ the conversion ratio and $d$ the predator mortality rate.  Here we have assumed for the prey, $\xi_x \sim \mathcal{N(0, \sigma_x)}$, and similarly for the predator driven by $\xi_y$. Figure 1A shows examples for noise $\sigma_x = \sigma_y = 10^{-5}$, while Figure 1B shows $\sigma_x = \sigma_y = 0.01$.

<!-- Figure 2: Quasi-cycles -->
```{r quasicycles, fig.width=7, fig.cap="Panel A: under negligible environmental noise, the introduction of a carrying capacity damps Lotka-Volterra oscillations which eventually settle to a stable state.  Panel B: small noise induces quasi-cycles in the same model: sustained long-term oscillations predominately around the natural frequency of the original damped oscillations. Parameters for models shown are r = .1, K = 5,  b = .1, c = .1, d = .1, N = 800; code in Appendix A."}
read_csv("appendixA/quasicycles.csv") %>%
  gather(species, population, -t, -sigma) %>%
  mutate(sigma = as.factor(sigma)) %>%
  mutate(sigma = recode(sigma, "1e-05" = "A.", "0.01" = "B.")) %>%
  rename(time = t) %>%
  ggplot(aes(time, population, col = species)) + geom_line() + 
  facet_wrap(~ sigma, ncol=2) + scale_color_ptol() #+ 
#  labs(x="Time", 
#       y="Population density",
#       title="Noise induces quasi-cycles",
#       caption="Left panel: neglible  noise results only in damped oscillations.\nRight panel: small noise creates sustained oscillations.")
```

@@

## Stochastic switching

We turn from stochastic resonance to oscillations of a very different sort, which do not resemble any deterministic process, as seen in Figure 3.

<!--Figure 3: Noisy switch -->
```{r noisy_switch, fig.width=7, fig.cap="Stochastic switiching in May's model of alternative stable state dynamics.  r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 10000"}

panel1 <- read_csv("appendixA/noisy_switch.csv")  %>%
  ggplot(aes(t,x)) + 
  geom_line(lwd=.1, col=colours[[1]]) +
  labs(subtitle = bquote(bold(A.)))

p <- list(r = .5, K = 2, Q = 5, H = .38, sigma = .04, a = 0.245, N = 1e4)
f <- function(x) x * p$r * (1 - x / p$K)
g <- function(x) p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)
x <- seq(0,2, length.out = 100)

panel2 <- data.frame(x, f = f(x), g = g(x)) %>%
  ggplot(aes(x, y)) + 
  geom_line(aes(y = f), col=colours[1], lwd=1) +
  geom_line(aes(y = g), col=colours[2], lwd=1) + 
  labs(x = bquote(X[t]), y = bquote(X[t+1]), subtitle= bquote(bold(B.)))

library(gtable)
library(grid) # low-level grid functions are required
g1 <- ggplotGrob(panel1)
g2 <- ggplotGrob(panel2)
g <- cbind(g1, g2, size="first") # stack the two plots
#g$widths <- unit.pmax(g1$widths, g2$widths) # use the largest widths
grid.newpage()
grid.draw(g)
```

This pattern is characterized by long stretches during which fluctuations in the state variable remain concentrated around a high average value followed by sudden, rapid transistions into fluctuations concentrated around a lower value.  This behavior arises through the interaction of stochasticity and alternative stable state dynamics.  The simulation shown in Figure 3 is based on an herbivore consumption model first proposed by @May1977 as one of the early examples of the potential for alternative stable state dynamics, to which I have simply introduced an additional environmental noise term:

$$X_{t+1} = X_t + \underbrace{X_t r \left(1 -\frac{X_t}{K} \right)}_{\textrm{Vegitation growth}}  - \underbrace{\frac{a X_t ^ Q}{X_t^ Q + H ^ Q}}_{\textrm{Vegitation consumption}} + \xi_t,$$
where $\xi_t$ is normally distributed environmental variation with mean zero and standard deviation $\sigma$.

@@

[@Boettiger2012]

[@Drake2013]

@@

<!-- Figure 4: Early warning signals -->
```{r tipping, fig.width=7, fig.height=7}
tipping <- read_csv("appendixA/tipping.csv")

p1 <- tipping %>%
  select(-x) %>%
  rename("A. autocorrelation" = autocorrelation, "B. variance" = variance) %>%
  gather(series, value, -t, -tip_time) %>%
  ggplot(aes(t, value)) + geom_line(col=colours[[1]]) + 
  geom_vline(aes(xintercept = tip_time), col=colours[[2]], lty=2) +
  facet_wrap(~series, ncol = 2, scales="free_y")  +
  theme_ipsum_rc() 
p2 <- tipping %>% 
  select(t, population = x, tip_time) %>%
  ggplot(aes(t, population)) + geom_line(col=colours[[1]]) + 
  geom_vline(aes(xintercept = tip_time), col=colours[[2]], lty=2) + 
  #scale_y_log10()  + 
  theme_ipsum_rc() + 
  labs(subtitle = "C. Yeast population size", 
       caption = "Vertical red dashed line indicates tipping point location")

library(cowplot)
ggdraw() +
  draw_plot(p1, x = 0, y = .5, width = 1, height = 0.5) +
  draw_plot(p2, x = 0, y = 0, width = 1, height = 0.6)
```
